English Vocabulary Learning Pattern

### 2024-11-23 paper/paper.qmd
Minor updats in scripts.

- High Vocabulary: The outcome variable, high_vocabulary, is a binary indicator that takes the value of 1 if the average production and comprehension score (prod_comp_mean) exceeds 350 and 0 otherwise. This threshold was selected to represent children with relatively advanced vocabulary skills, determined through exploratory data analysis.
- Scaled Age (age_scaled): This continuous variable represents the child’s age, standardized to ensure the model coefficients reflect changes per standard deviation in age. Standardization improves numerical stability and aids interpretability.
- Norming Status (is_norming): A binary indicator denoting whether a child is part of the norming dataset (TRUE) or not (FALSE). This variable accounts for potential differences in data collection or assessment protocols.
- Broad Category (broad_category): A categorical variable grouping words into four broad linguistic categories: nouns, verbs, adjectives, and function_words. The reference category is function_words.

The model takes the form:

@@

Where:
- $p_i$ represents the probability that person i has a high vocabulary
- $\beta_0$ is the intercept, capturing the baseline log-odds when all predictors are at their reference or mean levels
- $\beta_1$: Effect of age (standardized)
- $\beta_2$: The effect of whether the individual belongs to the norming group
- $\beta_3$, $\beta_4$, $\beta_5$, etc.: The effects of being in the respective broad word categories (nouns, function words, or verbs), compared to the reference category (likely “adjectives”).

@@

The model was trained on 80% of the dataset, with the remaining 20% reserved for testing. Model performance was assessed using confusion matrices and overall accuracy. Further evaluation included the analysis of residual deviance, AIC, and interpretation of individual coefficients.

@@

## Average Production by Broad Category and Age

@@

analysis_data %>%
  group_by(broad_category, age) %>%
  summarize(avg_production = mean(production, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = age, y = avg_production, fill = broad_category)) +
  geom_col(position = "stack") +  # Use stack for stacked bars
  labs(
    title = "Average Production by Broad Category and Age",
    x = "Age (months)",
    y = "Average Production Score",
    fill = "Broad Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11))
```

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: Checking the convergence of the MCMC algorithm

@@

>>>>

- High Vocabulary: A binary indicator where 1 represents a high vocabulary score (combined comprehension and production > 350), and 0 otherwise.

- Scaled Age (age_scaled): The child’s age, standardized to reflect changes per standard deviation. Standardization aids in interpretability and ensures numerical stability.

- Norming Status (is_norming): A binary indicator denoting whether a child is part of the norming dataset (TRUE) or not (FALSE). This variable accounts for potential differences in data collection or assessment protocols.

- Broad Category (broad_category): A categorical variable grouping words into lexical categories, such as adjectives, verbs, and nouns. The reference category for comparison is Function Words.

The model is specified as:

@@

Where:
- $p_i$ represents the probability that child  i  has a high vocabulary score.
- $\beta_0$ is the intercept, capturing the baseline log-odds when all predictors are at their reference or mean levels
- $\beta_1$: Effect of age (standardized)
- $\beta_2$: The effect of whether the individual belongs to the norming group
- $\beta_3$, $\beta_4$, $\beta_5$, etc.: The effects of being in the respective broad word categories (nouns, function words, or verbs), compared to the reference category (likely “adjectives”).

@@

The model was trained using robust statistical principles to ensure reliability:

	•	The split of data into training and testing subsets facilitates evaluation of the model’s predictive power on unseen data.
	•	Standardizing continuous predictors such as age improves interpretability and comparability across variables.
	•	Saving preprocessed datasets and the trained model ensures reproducibility and traceability of results.

@@

## Average Production by Broad Category

@@

# Filter the dataset for Adjectives and Objects
filtered_data <- analysis_data %>%
  filter(broad_category %in% c("Adjectives", "Objects"))

# Select 100 random unique child IDs
set.seed(123)  # Ensure reproducibility
sampled_child_ids <- filtered_data %>%
  distinct(child_id) %>%
  slice_sample(n = 100) %>%
  pull(child_id)

# Filter data for these 100 children
filtered_data_sampled <- filtered_data %>%
  filter(child_id %in% sampled_child_ids)

# Create a line plot for each child's production trend
ggplot(filtered_data_sampled, aes(x = age, y = production, color = as.factor(child_id), group = child_id)) +
  geom_line(alpha = 0.8, size = 1) +  # Line for each child
  facet_wrap(~ broad_category, scales = "free_y") +  # Separate facets for Adjectives and Objects
  labs(
    title = "Production Trends by Child Across Categories (Sampled 100 Children)",
    subtitle = "Adjectives and Objects",
    x = "Age (Months)",
    y = "Production Score",
    color = "Child ID"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 14),
    legend.position = "none",  # Hide legend to avoid clutter
    strip.text = element_text(size = 12, face = "bold")  # Customize facet labels
  )
```

@@

```{r}
library(ggplot2)

# Add predicted probabilities and logit to the dataset
analysis_data_train <- analysis_data_train %>%
  mutate(predicted_prob = predict(logistic_model, type = "response"),
         logit = log(predicted_prob / (1 - predicted_prob)))

# Plot logit against continuous variables (e.g., age_scaled)
ggplot(analysis_data_train, aes(x = age_scaled, y = logit)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(title = "Logit vs. Age (Scaled)",
       x = "Age (Scaled)",
       y = "Logit (log-odds)") +
  theme_minimal()
```

### Accuracy
```{r}
# Predicted probabilities and classes
predicted_classes <- ifelse(fitted(logistic_model) > 0.5, 1, 0)
confusion_matrix <- table(Predicted = predicted_classes, Actual = analysis_data_train$high_vocabulary)

# Print confusion matrix and accuracy
print(confusion_matrix)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 3)))
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-convergencecheck
#| fig-cap: "Checking the convergence of the Markov chain Monte Carlo (MCMC) algorithm for hit song model"
#| fig-subcap: ["Trace plot for model", "Rhat plot for model"]
#| layout-ncol: 2

# Creates trace plot
plot(predicted_prob, "trace") + theme_minimal()
# Creates rhat plot
plot(predicted_prob, "rhat") + theme_minimal()

@@

### Binned Residual Plot
```{r}
install.packages("arm")
library(arm)

binnedplot(fitted(logistic_model), residuals(logistic_model, type = "response"),
           xlab = "Fitted Probabilities", ylab = "Average Residual",
           main = "Binned Residual Plot")
```
